---
title: "S10 in class activity"
author: "Chenxin Li"
date: "2026-06-25"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 
In this lesson we will talk about:

1. How to do a Principal Component Analysis (PCA) in R.
2. How to interpret and explore the results of PCA. 

# Packages
```{r}
library(tidyverse)
library(RColorBrewer)
library(readxl)
```

PCA itself does not require any external packages. 
`tidyverse` and `RColorBrewer` is only for graphing. 

# Visualizing high dimensional data 
Let’s start with something simple and work our way up there. 
Let's load the penguin dataset from the Data folder as an example. 
```{r}
penguins <- read_csv("../Data/S4_example1.csv")
head(penguins)
dim(penguins)
```
We have 344 observations at 8 variables. 
If we want to visualize 2 variables, we can make a simple x-y scatter plot. 
```{r}
penguins %>% 
  ggplot(aes(x = bill_len, y = bill_dep)) +
  geom_point(aes(color = species), size = 3)+
  scale_color_manual(values = brewer.pal(n = 8, "Set2")) +
  labs(x = "bill length",
       y = "bill depth") +
  theme_classic()
```
When we have high dimensional data, 
we will need to perform dimension reduction using PCA to flatten the data into 2-D. 
The main idea of PCA is that it finds correlated variables
and collapses them into a shared axis called a principal component (PC).   
When R performs PCA, the first PC (PC1) will span the most variation in the dataset,
followed by PC2, PC3, and so on. 

# Running PCA 
To run PCA in R, you want to have the following:  

1. A numeric matrix where rows are samples and columns are variables. 
2. log-transform or z-score for the input matrix is recommended, so that the PCs won't be dominated by variables with higher numeric values. 
3. Row names of the matrix to identify which sample is which for downstream analysis. 
4. For graphing and downstream analysis, descriptions for each sample will be helpful (metadata). 

Let's try an example. 
```{r}
bean_heat <- read_excel("../Data/S10_example1.xlsx")
head(bean_heat)
```

Data were taken from [this paper](https://www.nature.com/articles/s41467-021-22858-x). In this study, the authors looked at gene expression of tepary bean leaves under control and heat stress. They sampled 5 time points during the stress treatment. One of the goals of the experiment was to understand gene regulation in response to heat stress at early stages of stress.

We can analyze the experimental design with a table. 

|   source     | factor | levels |
|:------------:|:------:|:------:|
|  heat stress | treat  |   2    |
|  time point  | treat  |   5    |
|  libraries   | OU, EU |   30   | 

This is a multifactorial experiment. 
```{r}
dim(bean_heat)
```
It looks like each row is a gene, and each column is a variable. 
The experiment detected 27541 expressed genes across 30 libraries. 
The values in the table are normalized expression (FPKM, according to the authors). 
Also, according to the authors, the library "C_12hr_B1" failed. We will go ahead and remove it. 
I also found that some genes were repeated in the table.
There is no explanation on why that is. 
I removed the duplicated gene names by keeping their first entry. 
As you can see, real data can be messy. 

## log transform and z score 
First, I will use tidyverse functions to turn this table into a tidy form. 
Then, I will log transform the expression values. 
This is recommended so that the PCA won't be dominated by the most highly expressed genes. 
```{r}
bean_heat_long <- bean_heat %>% 
  distinct(LocusName, .keep_all = T) %>% 
  dplyr::select(-C_12hr_B1) %>% 
  pivot_longer(cols = !LocusName, names_to = "library", values_to = "FPKM") %>% 
  mutate(FPKM = case_when(
    is.na(FPKM) ~ 0,
    T ~ FPKM
  )) %>% 
  mutate(logFPKM = log10(FPKM + 1)) 

head(bean_heat_long)
```

## Long to wide 
PCA requires a matrix, so we have to go back to wide format again. 
We will need libraries as rows and genes as columns, and fill the table with logFPKM values.  
We will also need row names to keep track of which library is which after we do PCA. 
```{r}
bean_heat_for_PCA <- bean_heat_long %>% 
  dplyr::select(LocusName, library, logFPKM) %>% 
  pivot_wider(names_from = LocusName, values_from = logFPKM) %>% 
  as.data.frame()

row.names(bean_heat_for_PCA) <- bean_heat_for_PCA$library

head(bean_heat_for_PCA)
```

## Run PCA
To run PCA, call `prcomp()` in R. 
We just need to feed `bean_heat_for_PCA` (not including the `library` column) into `prcomp()`.
```{r}
my_pca <- prcomp(bean_heat_for_PCA[, -1])
```

# How much variance does each PC explain? 
The first diagnostic of PCA is to check how much variance each PC explains. 
See the code below to understand how to do it. 
```{r}
pc_importance <- as.data.frame(t(summary(my_pca)$importance))
pc_importance
```
You can see that PC1 explains 48% of the variance, followed by PC2, which explains the 13%. 
Between the first 2 PCs, we explained more than half of the variances.
You can also plot it. 
```{r}
pc_importance %>% 
  ggplot(aes(x = 1:29, y = `Proportion of Variance`)) +
  geom_bar(stat = "identity") +
  labs(x = "PC 1-29") +
  theme_classic()

pc_importance %>% 
  ggplot(aes(x = 1:29, y = `Cumulative Proportion`)) +
  geom_bar(stat = "identity") +
  labs(x = "PC 1-29") +
  theme_classic()
```
I've seen experiments where the first 2 PCs explained >90% of variances. 
But overall, this is what you should typically expect. 

# Graphing PCA results
The results from PCA are stored as a list. One of the objects in the list is called `x`.
`x` stores the coordinate of each sample in the PC space,
where each row is a sample, and each column is a PC. 


## Metadata
To facilitate graphing, we need metadata to explain which library is which. 
According to the naming convention of the authors: 

* `C` vs `S` in library names stands for control vs stressed.
* Data were collected at 1, 3, 6, 12, and 24 hrs post treatment.
* Library names ending with B1, B2, or B3 correspond to the 3 bio-reps. 

We can parse that out:
```{r}
Metadata <- data.frame(
  library = colnames(bean_heat)[-1]
) %>% 
  mutate(treatment = case_when(
    str_detect(library, "C_") ~ "control",
    str_detect(library, "S_") ~ "heat"
  )) %>% 
  mutate(time_point = case_when(
    str_detect(library, "_1hr_") ~ "1",
    str_detect(library, "_3hr_") ~ "3",
    str_detect(library, "_6hr_") ~ "6",
    str_detect(library, "_12hr_") ~ "12",
    str_detect(library, "_24hr_") ~ "24"
  )) %>% 
  mutate(time_point = factor(time_point, levels = c(1, 3, 6, 12, 24))) %>% 
  filter(str_detect(library,"C_12hr_B1") == F)

dim(Metadata)
```
Now we simply need to join the metadata to `x`. 
```{r}
PCA_coord <- my_pca$x %>% 
  as.data.frame() %>% 
  mutate(library = row.names(.)) %>% 
  full_join(Metadata, by = "library")

head(PCA_coord)
```

## Plot PCA
```{r}
PCA_coord %>% 
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes(fill = time_point), color = "grey20", shape = 21, size = 3, alpha = 0.8) +
  scale_fill_manual(values = brewer.pal(n = 5, "PiYG")) +
  labs(x = paste("PC1 (", pc_importance[1, 2] %>% signif(3)*100, "% of Variance)", sep = ""), 
       y = paste("PC2 (", pc_importance[2, 2] %>% signif(3)*100, "% of Variance)", "  ", sep = ""),
       fill = "time point") +  
  theme_bw() +
  theme(
    text = element_text(size= 14),
    axis.text = element_text(color = "black")
  )
```

```{r}
PCA_coord %>% 
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(aes(fill = treatment), color = "grey20", shape = 21, size = 3, alpha = 0.8) +
  scale_fill_manual(values = c("grey70", "tomato1")) +
  labs(x = paste("PC1 (", pc_importance[1, 2] %>% signif(3)*100, "% of Variance)", sep = ""), 
       y = paste("PC2 (", pc_importance[2, 2] %>% signif(3)*100, "% of Variance)", "  ", sep = ""),
       fill = "treatment") +  
  theme_bw() +
  theme(
    text = element_text(size= 14),
    axis.text = element_text(color = "black")
  )
```

It might not make sense in the 1st glance, but time is circular! 24hr is very close to 1hr in the diurnal sense. 
It seems PC1 separates midnight (time point 24) from noon (time point 12), 
and PC2 is driven by the difference between time point 6 and the rest of the time points. 

```{r}
PCA_coord %>% 
  ggplot(aes(x = PC2, y = PC3)) +
  geom_point(aes(fill = treatment), shape = 21,size = 3, alpha = 0.8, color = "grey20") +
  scale_fill_manual(values = c("grey70", "tomato1")) +
  labs(x = paste("PC2 (", pc_importance[2, 2] %>% signif(3)*100, "% of Variance)", sep = ""), 
       y = paste("PC3 (", pc_importance[3, 2] %>% signif(3)*100, "% of Variance)", "  ", sep = ""),
       fill = "treatment") +  
  theme_bw() +
  theme(
    text = element_text(size= 14),
    axis.text = element_text(color = "black")
  )
```
If we plot the next PC, which is PC3 against PC2, 
we will find that the control and heat treatments are separated along PC3. 

```{r}
PCA_coord %>% 
  ggplot(aes(x = PC3, y = PC4)) +
  geom_point(aes(fill = treatment), shape = 21,size = 3, alpha = 0.8, color = "grey20") +
  scale_fill_manual(values = c("grey70", "tomato1")) +
  labs(x = paste("PC3 (", pc_importance[3, 2] %>% signif(3)*100, "% of Variance)", sep = ""), 
       y = paste("PC4 (", pc_importance[4, 2] %>% signif(3)*100, "% of Variance)", "  ", sep = ""),
       fill = "treatment") +  
  theme_bw() +
  theme(
    text = element_text(size= 14),
    axis.text = element_text(color = "black")
  )
```
If we plot the next PC, which is PC4 against PC3, 
we will find that PC4 separates a single library from the rest. 
Since the first 3 PC together explained most of the variance for this experiment, 
we are probably just looking at the noise in each individual samples for PC4 and beyond. 

# Explore which variable drives separation along PCs
One of the results of PCA is stored as the `rotation` matrix. 
This is also called the "loading" on each PC. 
In this matrix, each row is a variable, which is a gene.
Each column is a PC. 
A positive value means that when a gene is higher expressed, 
it moves a sample further along that PC. 
A negative value means that when a gene is higher expressed, 
it moves a sample to the opposite direction of the PC. 

Let's say we are interested in heat stress, 
and we know heat vs control is separated along PC3. 
We can check which genes have the highest (most positive or most negative) loading for PC3. 
```{r}
heat_candidate_genes <- my_pca$rotation %>% 
  as.data.frame() %>% 
  mutate(LocusName = row.names(.)) %>% 
  pivot_longer(cols = !LocusName, names_to = "PC", values_to = "loading") %>% 
  filter(PC == "PC3") %>% 
  arrange()

heat_candidate_genes %>% 
  ggplot(aes(x = loading)) +
  geom_histogram(bins = 100) +
  theme_classic()
```
You can see most of the genes have 0 loading for PC3. 
However, there are a small number of genes that are quite far out in the distribution at either extreme. 
Let's take the top 3 and bottom 3 genes and check their expression. 
```{r}
heat_candidate_genes %>% 
  slice_max(order_by = loading, n = 3) %>% 
  inner_join(bean_heat_long, by = "LocusName") %>% 
  inner_join(Metadata, by = "library") %>%  
  ggplot(aes(x = time_point, y = logFPKM)) +
  facet_wrap(~LocusName)+
  ggbeeswarm::geom_quasirandom(aes(color = treatment)) +
  stat_summary(geom = "line", 
               fun.data = mean_se, aes(group = treatment, color = treatment), linewidth = 1.2)+
  stat_summary(geom = "linerange", 
               fun.data = mean_se, aes(group = treatment, color = treatment))+
  scale_color_manual(values = c("grey70", "tomato1")) +
  labs(x = "time point")+
  theme_classic()
```
The top 3 genes with the highest loading on PC3 are all higher in the control relative to the heat treatment.
This is as expected, since control samples are higher along PC3 relative to the heat sample. 

```{r}
heat_candidate_genes %>% 
  slice_min(order_by = loading, n = 3) %>% 
  inner_join(bean_heat_long, by = "LocusName") %>% 
  inner_join(Metadata, by = "library") %>%  
  ggplot(aes(x = time_point, y = logFPKM)) +
  facet_wrap(~LocusName)+
  ggbeeswarm::geom_quasirandom(aes(color = treatment)) +
  stat_summary(geom = "line", 
               fun.data = mean_se, aes(group = treatment, color = treatment), linewidth = 1.2)+
  stat_summary(geom = "linerange", 
               fun.data = mean_se, aes(group = treatment, color = treatment))+
  scale_color_manual(values = c("grey70", "tomato1")) +
  labs(x = "time point")+
  theme_classic()
```
The top 3 genes with the most negative loading on PC3 are all higher in the heat treatment relative to the control.
This is as expected, since control samples are higher along PC3 relative to the heat sample. 
A negative loading will move the samples to a lower PC3 value. 

# In class activity 
## Q1
Load "S10_activity1.csv" from the Data folder. 
This is a transcriptome dataset with 18 samples from 6 tissues.  
In the table, each row is a gene, and each column is an RNA-seq library. 
The 6 tissue types are: 

1. immature top leaf
2. middle leaf
3. mature bottom leaf
4. primary root
5. secondary root 
6. immature top stem 

Each tissue type has 3 replicates, labelled as R1, R2, and R3, respectively. 

1. Run a PCA on this dataset.
2. How much variance does each PC explain? 
3. Parse the sample IDs to produce a metadata table. 
4. Make a PCA plot. 
5. Explore the PCA results. 








