---
title: "Session 1 in-class activity"
author: "Chenxin Li"
date: "2026-06-21"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This is a header 
## This is a sub-header 

If you change the output to html_notebook, 
it will automatically render an HTML file on save.   
You should be able to find the HTML file in the same folder as your .Rmd script. 

You can write any plain text you want here.   
You can _italicize_ or __bold__ text if you want.  
You can ~~cross out~~ text as well.  
You can start a new line using two spaces at the end of a line.  

You can start a list like this: 

* First item of list 
* Second item of list 
* Remember to leave an empty line before the first bullet point 

Or like this: 

1. First item
2. Second item 
3. If you want numbered list

This RMarkdown is for you to familiarize with RStudio, 
and learn how to load packages into your session.   
At the start of your RMarkdown, you want to have a brief 
introduction section to describe what this script or notebook is doing. 

# Packages 
After the introduction section, you will start a new section to load your packages. 
To do that, you will need to insert a code chunk. 
The short cut is `ctrl-alt-i` on PC or `cmd-opt-i` on Mac. 
Now you have a code chunk.  
```{r}
# This is where R will run your code. 
# If you need to write comments here you will need to start the line with a # sign. 
```

To load a package, you use `library()` function and specify the package name 
inside the parentheses.   
To run a code chunk, click the green triangle at the upper right of the code chunk.  
Or you can do `ctrl-enter` on a PC or `cmd-enter` on a Mac to run line-by-line. 

```{r}
library(tidyverse) 
```

You might get some warning messages regarding the version of R.  
As long as there is no error messages, you are all good. 
 
The tidyverse is a collection of packages for data science and visualization.  
These packages use a similar style of syntax and design principal.  
While it is not the fastest or keystroke efficient, we will adopt these packages 
due to how intuitive they are. 

# How to get help 
Say you want to know how to use the `head()` function. 
You type ?head in the console.  
RStudio will pull up the help page in the lower right corner.  
It should say "returns the first or last parts of objects" or something similar. 

# Formatting a table in markdown 
To render a table in html using markdown, you do the following: 

|  Header1  |  Header1  | Header3|
|:---------:|:---------:|:------:|
|     1,1   |   1,2     | 1,3    |
|     2,1   |   2,2     | 2,2    |

You can check in the html file to see if the table has been rendered correctly. 
The alignment of text is controlled by `|:---:|` syntax.

* `|:---:|` means aligned at the center. 
* `|:---|` means aligned to the left (also the default).  
* `|---:|` means aligned to the right. 

# In-class activity 
Start a new Rmd file. 
Change output from html_document to html_notebook

# Q1 
Briefly introduce yourself. 
	* Which degree program are you in? 
	* What year are you in? 
	* If known, which lab are you part of? Very briefly describe what you are working on. 
	* What do you want to learn in 302? 

Feel free to use formatting options such as italic, bold, lists, and headers.  

# Q2
Consider this hypothetical experiment:  
You were interested in the effect of mechanical stimulation on plant defense. 
You had 2 trays of plants with 30 plants in each tray. For one tray you used a paint brush to touch the plant 3 times a day. 
For the other tray you didn't do anything (the control tray). 
After 3 months, you randomly sampled leaves from plants in each tray. 
You took 3 tubes of samples per tray and each tube has 3 leaves.  
You extracted RNA from each tube and sent the RNA for RNA-seq. 

1. Analyze this experimental design using a table. What was the dependent variable? 
What were the treatment and sources of variation? 
What were the experimental and observational units? 
How many levels were there per treatment, and what was the sample size? 
Hint: How many tubes of RNA are you sending out? 

2. If it costs $5 per sample for RNA extraction, and 
$200 per sample for library prep + sequencing, how much will this experiment cost?  

# Q3 
You are working on chemical manufacturing using plant tissue culture. 
You believe a substance (compound X) can stimulate flavor production in wasabi tissue culture. 
You prepared 4 different growth media with concentrations of X at 0, 1, 10, 100 mM.  
Growth media were poured into 10 petri dishes per concentration. 
20 tissue pieces of wasabi were grown in each petri dish. 
2 weeks after culturing, you took 3 samples per treatment by taking 5
tissue pieces per sample, where tissue pieces were randomly chosen across petri dishes per treatment. 
Each sample is collected into a test tube.
For each test tube, you ground the tissue into fine powder, extracted metabolites,  
and brought them to the Mass Spec Core for metabolite profiling. 
Using the mass spec result, you quantified two main flavor compounds from wasabi: sinigrin and 6-(Methylsulfinyl)hexylglucosinolate.

1. Analyze this experimental design using a table. What were the dependent variables?
What were the treatments and sources of variation? 
What were the experimental and observational units? 
How many levels were there per treatment, and what was the sample size? 
Hint: How many samples of metabolites do you need to run on the mass spec? 

2. Let's say it takes 10 min per sample to run on the mass spec, 
and the mass spec core charges $56 per hour.  
Your PI told you that you can only spend $300 or less at the Mass Spec Core. 
Will you have enough funds for this experiment?  
 

