---
title: "Session 2 in-class activity"
author: "Chenxin Li"
date: "2026-06-22"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 
In this lesson, we will go over how to: 

* Calculate summary statistics using tidyverse functions. 
* Use probability distribution functions in R. 

# Package
We will need tidyverse. 
```{r}
library(tidyverse)
```
# Summary statistics  
Tidyverse has an easy syntax to calculate summary statistics, such as mean and median from data frames using `summarise()`. 
Summary statistics that can be plugged into `summarise()` includes:

* average: `mean()`
* median: `median()`
* variance: `var()`
* standard deviation: `sd()` 
* interquartile range: `IQR()`
* sample size: `n()` 

And many others. 

Let's try it out on a hypothetical experiment. 
```{r}
# You can ignore this chunk. 
trt <- rnorm(n = 10, mean = 10, sd = 1)
ctrl <- rnorm(n = 10, mean = 6, sd = 1)
location <- c(rep("A", 5), rep("B", 5))

example_experiment <- data.frame(
  ctrl,
  trt,
  location
) %>% 
  pivot_longer(names_to = "treatment", values_to = "response", cols = 1:2)

example_experiment
```

In this experiment, we have two treatments, control (ctrl) and treated (trt).  
The experiment was repeated at two locations, A or B.  
At each location, 10 samples were collected. 
The dependent variable is recorded in the "response" column. 

If we want to find the mean, median, var, sd, IQR, and n for the whole experiment, 
we will do: 
```{r}
example_experiment_s <- example_experiment %>% 
  summarise(
    mean = mean(response), 
    median = median(response),
    var = var(response),
    sd = sd(response),
    IQR = IQR(response),
    n = n()
  ) 

example_experiment_s
```
The resulting data frame has 1 row and 6 columns, with each column being a summary statistic. 

## Summary statistics by group 
It will be more informative to calculate summary statistics by group (ctrl vs. trt)  
so that we can understand if the treatment had an effect on the dependent variable. 
To do that, we simply need to perform `group_by(treatment)` before running `summarise()`.
```{r}
example_experiment_s_by_group <- example_experiment %>% 
  group_by(treatment) %>% 
  summarise(
    mean = mean(response), 
    median = median(response),
    var = var(response),
    sd = sd(response),
    IQR = IQR(response),
    n = n()
  ) %>% 
  ungroup()

example_experiment_s_by_group
```


1.  `group_by()` selects which columns you want to summarize from.
    In this case you just want each treatment to be a group.
2.  `summarise()` allows you to call the statistical functions, such as mean, sd, and n.
3.  It's a good practice to `ungroup()` after you are done.
    Most of the time it doesn't matter, but it will create problems if you don't and you try to switch/modify grouping later.

Now you just get 2 rows and 4 columns with each row as a group. 

Say I want summary statistics for each group and for each location. Can I do that?
Yes, you just need to also include location in `group_by()`. 
```{r}
example_experiment_s_by_group_location <- example_experiment %>% 
  group_by(treatment, location) %>% 
  summarise(
    mean = mean(response), 
    median = median(response),
    var = var(response),
    sd = sd(response),
    IQR = IQR(response),
    n = n()
  ) %>% 
  ungroup()

example_experiment_s_by_group_location
```

1.  This time you need to select group and treatment in `group_by()`.
    Because you want the mean for each group and each location.
2.  In `summarise()` you call the mean, median, sd, and n functions.
3.  It's a good practice to `ungroup()` after you are done.

Now each location by treatment combination is a row. 
Because we have 2 treatments and 2 locations, we should have 2*2 = 4 rows total.  
In this experiment, we can see that ctrl vs. trt seems different for both locations. 

I hope you can see that `group_by()` followed by `summarise()` is very convenient and powerful. 
It saves you from having to use for loops to calculate statistics for each group. 
Instead, `group_by()` automates that.

# Probability distributions in R 
## Simulate data using probability distributions. 
R has several probability distributions already built-in, 
such as uniform, normal, and t. 
Each distribution comes with a family of functions: r, p, q, d. 
For example: 

* `rnorm()` lets you simulate random numbers that are normally distributed. 
* `pnorm()` lets you calculate the probability (upper or lower tail) given a number. 
* `qnorm()` is the reverse of `pnorm()`, given probability find number. 
* `dnorm()` lets you find the height of the probability density function at a given value. 

If you need the uniform distribution, these functions will be: `runif()`, `punif()`, `qunif()`, and `dunif()` instead. 

For example, if you want to simulate 10 normally distributed numbers, 
with mean = 1, and sd = 2, it will be
```{r}
norm10 <- rnorm(n = 10, mean = 1, sd = 2)
norm10
```
You will get a vector of 10 values. 

Say you want to find what is the probability of being 2 sd or more away from the mean 
given a normal distribution. 

"2 sd or more away from the mean" is the same as a normal distribution of mean = 0, 
sd = 1, and being at a value of 2 or more. We will need the upper tail distribution here, 
so we will set `lower.tail = F` in `pnorm()`. 
And we need to multiply by 2 to account for both extremes. 
We can simply multiply 2 here because normal distribution is symmetrical. 
```{r}
pnorm(q = 2, mean = 0, sd = 1, lower.tail = F) * 2 
```
It would be 0.046 or 4.6%.  

## Demonstration of central limit theorem. 
To demonstrate the central limit theorem, let's sample (n=3) from a uniform distribution 20,000 times. 
For each iteration, we calculate the mean and see what distribution the means fall into. 
Let's say the uniform distribution ranges from -1 to 1, so the mean would be its midpoint, 0. 
```{r}
set.seed(666)
my_data <- replicate(expr = runif(n = 10 , min = -1, max = 1), n = 20000, simplify = T)  %>% 
  t() %>% 
  as.data.frame() %>% 
  cbind(iteration = 1:20000) %>% 
  pivot_longer(cols = !iteration, names_to = "name", values_to = "value") %>% 
  group_by(iteration) %>% 
  summarise(
    mean = mean(value)
  ) %>% 
  ungroup()

head(my_data)
```

What happened here: 

1. Drew 3 numbers from the uniform distribution using `runif(n = 10)`
2. Repeated that 20000 times using `replicate()` 
3. Transposed the resulting matrix with `t()` such that each row is an iteration 
4. Converted to data frame using `as.data.frame()`
5. Performed a wide-to-long transformation using `pivot_longer()` 
6. Calculated the mean for each iteration. 

```{r}
my_data %>% 
  ggplot(aes(x = mean)) +
  geom_histogram(bins = 100, aes(y = after_stat(density))) +
  geom_density() +
  theme_classic()
```
You can see this looks like a normal distribution 
even though the original distribution (uniform distribution) looks nothing like a normal distribution.  
*Take home message*: Sample means are normally distributed around population mean, 
even if the population is not normally distributed. 

## Demonstration of the law of large numbers. 
To demonstrate the law of large numbers, let's sample from a normal distribution with increasing sample sizes. Let's say the normal distribution has mean = 0, sd = 1. 

```{r}
my_data2 <- list(
  "3" = rnorm(n = 3, mean = 0, sd = 1),
  "5" = rnorm(n = 3, mean = 0, sd = 1),
  "10" = rnorm(n = 10, mean = 0, sd = 1),
  "20" = rnorm(n = 20, mean = 0, sd = 1),
  "30" = rnorm(n = 30, mean = 0, sd = 1),
  "50" = rnorm(n = 50, mean = 0, sd = 1),
  "100" = rnorm(n = 100, mean = 0, sd = 1),
  "200" = rnorm(n = 200, mean = 0, sd = 1),
  "300" = rnorm(n = 300, mean = 0, sd = 1),
  "500" = rnorm(n = 500, mean = 0, sd = 1),
  "1000" = rnorm(n = 1000, mean = 0, sd = 1),
  "2000" = rnorm(n = 2000, mean = 0, sd = 1)
  ) %>% 
  lapply(as.data.frame) %>% 
  bind_rows(.id = "id") %>% 
  group_by(id) %>% 
  summarise(mean = mean(`X[[i]]`)) %>% 
  ungroup() %>% 
  mutate(sample_size = as.numeric(id))

my_data2
```

```{r}
my_data2 %>% 
  ggplot(aes(x = sample_size, y = mean)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_line(color = "tomato1") +
  geom_point(size = 3) +
  labs(x = "sample size", y = "sample mean") +
  theme_classic()
```
As you can see with increasing sample size, the sample mean gets closer and closer to the population mean. 

# In class activity 
# Q1 
Download the `readxl` package if you haven't. 
Use the `read_excel()` function to load "S2_activity1.xlsx" from the Data folder. 
In this [paper](https://link.springer.com/article/10.1186/s12870-020-2243-7), 
researchers evaluated the regeneration potential (number of buds and shoots) of daylily 
across 19 different varieties, 3 hormone treatments, and 3 explant types. 

Calculate the mean, sd, and IQR for each variety, hormone treatment, and explant type combination using tidyverse functions. 
Hint: How many variables do you need in the `group_by()` function? 

# Q2 
Let's say the population case frequency of a hypothetical disease X is 0.1%. 
There are two hospitals in town. One of them sees 200 patients a day and the other sees 10 patients a day. 
One of the hospitals filed a report saying among the patients they saw, 3% were positive for disease X, which is a big deviation from the 0.1% population level.  
Assuming the population case level for X has not changed, 
which hospital (the bigger one or the smaller one) likely filed the report? Why is that?  
Hint: What does the law of large numbers predict? 

# Q3 
Simulate two data sets with 30 numbers from the normal distribution. 
Both of them have mean = 0. One of them has a sd of 1, and the other has a sd of 2. 
Which data set will have a higher IQR? 
