---
title: "Session 4 in class activity"
author: "Chenxin Li"
date: "2026-06-23"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 
If you want to compare the means of two samples, what do you do?
You use the t-test.

However, what if you have three or more samples?
This lesson goes over one-way ANOVA and Tukey tests, which is designed to compare multiple ( >= 2) samples.

1.  How to set up a linear model?
2.  What are the assumptions of linear model and how to check them?
3.  How to interpret an ANOVA table?
4.  How to set up a Tukey test?
5.  How to interpret the compact letter display? -What it means and what it doesn't mean.

# Packages
You will need to install three new packages:

* `emmeans` is for pairwise means comparisons across groups 
* `multcomp` is for multiple testing correction 
* `multcompView` is for compact letter display for means separation 

We will go over how each of these is used in this lesson. 

```{r}
library(tidyverse)
library(emmeans)
library(multcomp)
library(multcompView)
```

# The t-test
The t-test works if you only have two groups to compare. 
For example: 
```{r}
group1 <- rnorm(n = 3, mean = 0, sd = 1)
group2 <- rnorm(n = 4, mean = 5, sd = 2)

t.test(group1, group2)
```
The function is `t.test()`. 
The default behavior in R assumes the variance is not equal between groups. 
The only assumption is both samples were drawn from a normal distribution. 

# Compare means across multiple groups.  
Let's start with an example. 
We will use the [penguins dataset](https://github.com/allisonhorst/palmerpenguins).
Ref: [Gorman KB, Williams TD, Fraser WR (2014) PLoS ONE 9(3): e90081.](https://doi.org/10.1371/journal.pone.0090081).
```{r}
penguins <- read_csv("../Data/S4_example1.csv")
head(penguins)
```

```{r}
dim(penguins)
```
Looks like we have 344 rows and 8 variables. 
Looking at the first 6 rows, this seems to be a tidy data frame. 
Each row is a penguin. Each column is a variable.   
There are three species of penguins in this table. They will be our independent variable.
As an example, we'll use `bill_len` (the length of the beak).  
We will compare bill length across the three species and find out who (if any) has the longest bill.
This means species will be the independent variable,
and bill length will be the dependent variable.  

## Data visualization 
But first, let's just quickly visualize the raw data to see what we are working with. 
```{r}
penguins %>% 
  ggplot(aes(x = species, y = bill_len)) +
  ggbeeswarm::geom_quasirandom() +
  labs(x = "species",
       y = "bill length") +
  theme_classic()
```
It looks like Adelie has the shortest bills. The other two species are somewhat comparable.  

## Summary stats
We can also check the summary statistics.
```{r}
length_s <- penguins %>% 
  filter(is.na(bill_len) == F) %>% 
  group_by(species) %>% 
  summarise(
    mean = mean(bill_len),
    var = var(bill_len),
    sd = sd(bill_len),
    n = n()
    ) %>% 
  ungroup()

length_s
```
# Setting up the linear model 
To perform an ANOVA, the dependent variable must be numeric,
and the independent variable must be categorical (or discrete factor). Can you check that? 
```{r}
head(penguins)
```

If you look at the first 6 rows of the data, 
`species` is recorded as a character variable, so it is categorical. 
`bill_len` is recorded as a numeric data, so we are all good to go. 

To perform an ANOVA in R, we set up the linear model first. 
Then we pass the model to ANOVA to test if any one mean is statistically different from the others. 
Don't forget to specify which data table you are using.  
```{r}
model_bill_len <- lm(bill_len ~ species, data = penguins)
```

# Linear model assumptions and how to check them 

1.  Independence: This means the measurement of one observation does not affect the other.
    In this example, we can only assume that's the case. 
    However, for example, if the same animal is measured twice,
    then it will violate the assumption of independence.
    If that is the case, you should first average across the repeated measures.

2.  Normality: the residues (errors) are normally distributed.
    Note: This is different from "the data is normally distributed".
    What this actually means is after the means are computed,
    the ERROR around the means (the residues) have to be normally distributed.
    The sample means don't have to be normally distributed.

You can check it with a normal qq plot using the following function: 
```{r}
plot(model_bill_len, which = 2)
```
This is called a normal qq plot, or normal quantile-quantile plot.
The quantiles of a normal distribution is plotted on x axis.
The quantiles of the model is plotted on the y axis.
If the residues are normally distributed, the dots will fall on a straight line across the diagonal.
In this model, the residues are normally distributed.

If you are not sure, you can simulate normal qq plots from a few normal distributions and check for yourself.

```{r}
rnorm(mean = 0, sd = 1, n = nrow(penguins)) %>% qqnorm() 
```

The code `rnorm()` generates normally distributed random numbers.
The `n = nrow(penguins)` argument generates the same number of data points as in the penguins data.
Every time you run the above chuck, the output will be different
because the numbers are randomly generated.
Run it several times to have an idea of what normally distributed data should look like.

3.  The last assumption is homoscedasticity: equal variance across groups.
    This may seem a bit abstract, but in fact it is very straightforward for one-way ANOVA.
    In this example, we have three groups, that is, three species.
    And equal variance would be the three species have roughly the same variance.
    Use the following function to check that: 

```{r}
plot(model_bill_len, which = c(1, 3)) 
```

The first graph is called residues vs fitted.
It is plotting the means of groups (in this case species) on x axis
and the spread of errors on the y axis.
If the variances are equal, you should see roughly the same amount of spread across the line.
It looks fine to me.

The second graph is called scale location.
Again x axis is still the means of groups,
but the y axis is now square root of absolute values of errors.
If the variances are equal, you should see roughly a flat line.
It looks more or less fine to me.

Another indication is that when you check the summary statistics, 
the variances are all within an order of magnitude of each other. 
The variance is not wildly different between groups. 

4. The last thing to check is outliers. 
In this case, when you plot the data, you saw there were no outliers.

Great! We are all clear to proceed with ANOVA.

# ANOVA table and its interpretation 
```{r}
anova(model_bill_len)
```

This is the output of an ANOVA. It's a table.
How do I read this?

In the applied sense, the most important number is the F value.
The F value is computed by mean sum of squares of dependent variable (species in this case)
over the mean sum of squares of residues, so 3597.2/8.8 = 410.

The F value can be understood as the variance between groups 
over the variances within group.
The null hypothesis is F = 1.
This means there is equal variation between groups and within groups.

In this example, we have F = 410, which is >> 1.
The p value, or Pr(>F) is the probability of finding F values more extreme than the observed F value,
if F = 1.
In this case, we have F = 410. If F was equal to 1, the probability of finding F >= 410 is < 2.2e-16.
This is very small.
Thus, we should reject the null hypothesis of F = 1.

This means the means of different species are different.

# How to do a Tukey test
ANOVA told us the means of different penguin species are different.
But it didn't tell us which one is the greatest.
To find that out, we need to use Tukey test.

```{r}
estimate_bill <- emmeans(model_bill_len, pairwise ~ species)
```

The `emmeans()` function means estimate marginal means.
The underlying math is actually pretty complex,
but for the applied sense, it can be understood as estimating the means of different groups.  
One thing to note is that marginal means differ from arithmetic means, 
as marginal means are based on linear models. 
With different linear models, marginal means can be different. 

In the `emmeans()` function, you will call which model the marginal means should be based on. 
You will call what type of comparisons we want to make,
and usually it is `pairwise`, meaning all pairwise contrasts.
Lastly, you will call which variable you want to compare across. In this case `species`.

After we estimate the means, we can compare the means.
```{r}
estimate_bill$contrasts
```

The is the output of Tukey tests and it is a table.

The contrast column should be self-explanatory: What is compared to what?
The estimate column is the difference of means.
The SE column is standard error of the differences of means.
The df column is degrees of freedom.

In the most applied sense, the most important numbers are the t.ratio.
t ratio is (the difference in mean) / SE.
For example, in row 1, t = -10.04 / 0.432 = -23.
The null hypothesis is t = 0, meaning no differences in mean.

The p.value means if t = 0, given the degrees of freedom,
what is the probability of getting a t.ratio more extreme than what's observed?

For example, in row 1, we have t = -23. If t were to be 0,
the probability of finding t <= -23 and more extreme is <0.0001.
This is very small, so we should reject the null hypothesis.

Keep in mind that Tukey tests give you the adjusted P values for multiple comparisons.
In this example, it stated "P value adjustment: tukey method for comparing a family of 3 estimates ".
What does this mean?

Because we have three species, we have three comparisons. They are:

1.  Adelie - Chinstrap
2.  Adelie - Gentoo
3.  Chinstrap - Gentoo

Under a p = 0.05 cutoff, if each comparison has a type I error rate of 0.05,
you have a total type I error rate of 0.05 * 3 = 0.15 in three tests.
Keep in mind that when you run a Tukey tests, R will take care of that,
and the p values are already adjusted,
so that the total type I error rate of your analysis is still 0.05.

So what do we learn from the table?
The means of the three species are all significantly different from each other.

## How to report the results of a Tukey test with compact letter display 
You can directly report the contrast table. 
However, another way is "compact letter display".
The compact letter display method is used in scientific publications a lot.
It's important that you know what it means and what it doesn't mean.

You will call the `cld` (stands for compact letter display) function, 
and put in the `emmeans` as the argument.
```{r}
cld(estimate_bill$emmeans, Letters = letters)
```
How do I interpret this output?
The emmean column is the estimated means.
The lower.CL and upper.CL are the confidence limits;
together they mark the 95% confidence interval of the mean.
Lastly, there is the ".group" column - how to interpret it?

If two rows share the same letter, it means the means of that two groups are not statistically different.
In this example, each species has their own unique letter,
so the means of the species are all significantly different from each other,
consistent with what the contrasts told us.

We can conclude that the species Chinstrap has the longest bill, and the species Adelie has the shortest.

As a practice, let's try a hypothetical example. In this cld, we have

| treatments | letter |
|:----------:|:------:|
|  Trt1      | a      |
|  Trt2      | ab     |
|  Trt3      |  b     |
|  Trt4      |   c    |

What does this mean?

It means:

* The means of Trt1 and Trt2 are not significantly different, because they both share the letter "a".
* The means of Trt2 and Trt3 are not significantly different, because they both share the letter "b.
* The means of Trt4, however, is significantly different from Treatments 1, 2 and 3, because it has the unique letter "c".

Be *VERY CAREFUL* when you interpret a cld output.
Two samples sharing the same letter does NOT mean their means are equal!
It only means their means are not significantly different in this experiment.
Their means might actually be different;
it's just the experiment didn't have enough power to detect that difference.

# In class activity 
## Q1 
Compare the flipper length `flipper_len` across 3 species of penguins in the `penguins` datast. 
Which species has the longest flipper? 

1. Visualize the data 
 
2. Find summary statistics: mean, var, sd, and n across species. 
Hint: Filter out missing data. 
 
3. Set up a linear model and check assumptions. 
 
4. Run ANOVA and interpret the ANOVA table. 
 
5. Run a Tukey test and interpret the contrast. 
 
6. Report CLD and interpret the output. 
 









