---
title: "Session 6 in class activity"
author: "Chenxin Li"
date: "2026-06-23"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

We went over how to perform one-way ANOVA to compare means of multiple treatments.
This time we'll cover a common experimental design, that is randomized block design. 

We'll go over

1.  What are blocking factors
2.  Why they may be useful 
3.  How to analyze an experiment that has blocking factors

# Packages
```{r}
library(tidyverse)
library(emmeans)
library(multcomp)
library(multcompView)
```

# What is a blocking factor? 
As an example, we'll use the R built-in dataset OrchardSpray.
This is an experiment with different spray treatments and compares their effect (the decrease of pests).
Ref: McNeil, D. R. (1977) Interactive Data Analysis. New York: Wiley.
```{r}
head(OrchardSprays)
```

It has 64 observations with 8 different treatments. 
`decrease` is decrease in pest count, which will be our depedent variable. 
To set up a linear model for ANOVA, it would just be `lm(decrease ~ treatment)`.  

However, there are two other columns: "rowpos" and "colpos".
They stand for row position and column position. These are blocking factors.
What exactly are they?

The best way to understand them is to plot out the experimental design in ggplot.
```{r}
OrchardSprays %>% 
  ggplot(aes(x = colpos, y = rowpos)) +
  geom_tile(color = "black", fill = NA) +
  geom_text(aes(label = treatment)) +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7, 8)) +
  scale_y_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7, 8)) +
  labs(x = "columns",
      y = "rows") +
  theme_classic()+
  coord_fixed()
```

This is the design of the experiment.
It is an 8x8 field. Each spot received a spray treatment (one of A-H),
and the treatments (A-H) were randomized across rows and columns.
We call the rows and columns blocking factors, in the sense that each row is a block, and each column is a block.

In reality, blocking factors can be as simple as a test tube rack. 

-   rack 1
    | a | b | b | a | a | b |

-   rack 2
    | a | b | b | a | a | b |

In this example, I have two racks; each rack has 6 test tubes.
Each rack has two treatments (a vs. b), and 3 replicates each.
Each rack can be used as a blocking factor.

In reality, blocking factors can be people as well.
Let's Say two students, Adel and Ben, needed to weigh 500 apples.
Adel weighed half of them, and Ben weighed the rest.
In this example, the person (Adel vs. Ben) can be used as a blocking factor.

# Why should one use blocking factors in an experiment?
Blocking factors allow you to control for unintended variations.

Let's use the 8x8 plot as an example.
Let's say there is a river near the right hand side of the orchard (to the right of column 8).
That means the right side of the orchard is wetter than the left side.
Thus, it make sense to have columns as a blocking factor, which controls for variation from left to right.

Let's say the upper side (row 8) is north, and the lower side (row 1) is south, and
this experiment happened in the northern hemisphere.
This means the south side of the orchard is getting more sunlight than the north side.
Thus, it make sense to have rows as a blocking factor as well, which controls for variation from top to bottom.

# How do I analyze an experiment with blocking factors?

In this experiment, we should do `lm(decrease ~ treatment)`, but what other things should I include in the model?

We can fill out an experimental design table. 

|  sources  |  factor   | levels |
|:---------:|:---------:|:------:|
|    row    |   block   |   8    |
|    col    |   block   |   8    |
| treatment | treatment |   8    |
|   spot    |  OU, EU   |   64   |

Rows and columns are blocking factors, and each has 8.
Treatment is the treatment, and there are 8 treatments.
Each spot is also a source of variation.

In this example, each spot is a replicate.
It is also the observational unit (OU). I.e., where the observations were made.
It is also the experimental unit (EU). I.e., the treatments were applied to each spot.

In many experiments, the observational unit is the experimental unit.
However, this is not always the case.  
The rule is to include all sources of variation in the model, other than the observational units.
Simply put, all factors other than OU.  
That makes `decrease ~ treatment + rowpos + colpos`

# Visualize the data
*WARNING*: `rowpos` and `colpos` are recorded as numeric variables in the table (1-8).
To use them as blocking factors, we need to change them into factors first. 
```{r}
OrchardSprays_new <- OrchardSprays %>% 
  mutate(row.F = as.factor(rowpos)) %>%  # mutate makes new columns based on existing columns  
  mutate(col.F = as.factor(colpos))
```

```{r}
OrchardSprays_new %>% 
  ggplot(aes(x = treatment, y = decrease)) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom()+
  labs(x = "treatments (A-H)",
       y = "decrease in pest count") +
  theme_classic()
```
# Summary statistics
```{r}
OrchardSprays_new %>% 
  group_by(treatment) %>% 
  summarise(mean = mean(decrease),
            var = var(decrease),
            sd = sd(decrease),
            n = n()) %>% 
  ungroup()
```
If you look at the variances, this is not ideal. Treatments A and B have a very low variance.
Treatments E and F have larger variances, but it doesn't look like there is a consistent trend on how variances change with means.
I don't think a log or logit transformation would be helpful.

(If you are interested, you can try log and logit transformation and compare the results.)
But for now we'll go with the original scale.

# Linear model
```{r}
model_spray <- lm(decrease ~ row.F + col.F + treatment, data = OrchardSprays_new)
```

Now let's check the assumptions of linear models 
## First off, normality.
```{r}
plot(model_spray, which = 2)
```

That's more or less okay.

## Second, equal variance across groups
```{r}
plot(model_spray, which = c(1, 3))
```
It looks rather okay to me. The spread is lower when the mean is lower, but overall it looks fine.  

## ANOVA
```{r}
anova(model_spray)
```

Treatment has a F value of 21, which rejects the null hypothesis of F = 1.
Row has a F value of 1.79, and 
Column has a F value of 1.35. The null hypothesis of F = 1 is not rejected (p > 0.05).
This means rows and columns do not explain much variation in this experiment.
This is actually a good thing. 
It means the row and column positions are not something you should worry about. 
Kind of a reassuring result, don't you think? 

## Tukey tests
Now we know there are differences among the means of different treatments.
Let's find out which treatment is the most effective (i.e highest decrease in pest).
Again we'll use the `emmeans` function. It will be the same thing as what we did last time.
```{r}
estimate_spray <- emmeans(model_spray, pairwise ~ treatment)
```
Since we have 8 treatments, we have 8 choose 2 = 28 different comparisons.
Let's pull out the cld (compact letter display).
```{r}
cld(estimate_spray$emmeans, Letters = letters)
```
What does this mean?
It means:

1.  The means of treatments ABCD are not significantly different from each other, because they all share the letter "a".
2.  The means of treatments DE are not significantly different from each other, because they both share the letter "b".
3.  The means of treatments EGFH are not significantly different from each other, because they all share the letter "c".

From this we can conclude that treatment GFH are the most effective treatments by Tukey tests.
Note that treatment H is numerically the highest,
but it is not detected as significantly different from treatments F,G and E.
How to reconcile that? There are two ways.

1.  The numerically greatest mean is due to pure chance. If you were to perform the experiment again, the mean of H may be lower next time.
2.  The mean of H is actually higher than the rest, but this experiment does not have enough power to detect that.
    If you were to re-run the experiment with larger sample sizes (say 16 instead of 8),
    you might detect a significant difference between H and the rest.

Based on the means, H is ~20 higher than the next highest (F and G). It's likely that interpretation 2) is correct.

Note that although A shares the letter "a" with B, and H shares the letter "c" with F,
*IN NO WAY* this proves mean(A) = mean(B), or mean(H) = mean(F).

Again, when you interpret P values and the letter grouping,
be thoughtful and careful on what they could mean, and what they don't mean.

# Including a continuous blocking factor (a covariate) 
A covariate is a continuous variable that can influence the response variable, 
but it is not the focus of the experiment. 
This can be used to control for unintended variations. 

Let's use the penguins dataset as an example. 
```{r}
penguins <- read_csv("../Data/S4_example1.csv")
head(penguins)
```

Let's say you want to compare the bill length between 3 species of penguins. 
You want to use body mass as a covariate to control for unintended variations due to body mass,
since penguins with a higher body mass may contribute to a longer bill. 
```{r}
model_body_mass <- lm(bill_len ~ body_mass + species, data = penguins)
anova(model_body_mass)
```
In this case, the F statistic for body mass is 623, which is even higher than the F statistic for species (399).
From this, you can interpret that body mass explains a higher amount of variance in bill length than species. 
If you want to see for yourself, you can plot it. 
```{r}
penguins %>% 
  ggplot(aes(x = body_mass, y = bill_len)) +
  geom_point(aes(color = species)) +
  geom_smooth(aes(color = species), se = F, method = "lm") +
  scale_color_manual(values = c("tomato1", "orange", "dodgerblue")) +
  labs(x = "body mass",
       y = "bill length") +
  theme_classic()
```
And you will see that even within each species, there is a good correlation between body mass and bill length. 

# In class activity
## Q1
Load "S6_activity1.csv" from the Data folder. 
The researcher was testing the effect of different pastures (P1 to P5).
on the amount of beta-carotene in eggs.
There were 4 fields (F1 - F4). The fields are the blocking factors.
In each field, hens were randomly assigned to be fed with one of the 5 pastures.
At the end of the experiment, eggs were collected and beta-carotene concentration was assayed.
(Data from UC Davis Plant Sciences course PLS205, Winter 2017) 

1. Visualize the data. 
2. Set up a linear model. 
3. Check assumptions of linear models. 
4. Run ANOVA and interpret the results. Do you think it was a good idea to include field as the blocking factor? 
5. Run a Tukey test and report results with cld. Which pasture treatment gave the best beta-carotene content?




