---
title: "Session 8 in class activity"
author: "Chenxin Li"
date: "2026-06-24"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 
In the previous lesson, we talked about randomized block design,
and how one can use blocks to control for unintended variations in an experiment.

This time we are going to talk about multifactorial designs. We will cover:

1.  What is a multifactorial design?
2.  Why would one design a multifactorial experiment?
3.  What is interaction between factors?
4.  How do you analyze a multifactorial experiment?

# Packages
```{r}
library(tidyverse)
library(emmeans)
library(multcomp)
library(emmeans)
```

# What is a multifactorial design?

A multifactorial design is an experiment that has multiple factors of independent variables.

We will use the R built-in dataset ToothGrowth as an example.
ToothGrowth is an experiment in which researchers looked at the effect of two supplements "VC" and "OJ" on the tooth length of guinea pigs.

Each animal was given a supplement (either OJ or VC);
and the animal can receive a high dose (2), mid dose (1) or low dose (0.5).
At the end of the experiment, the longest tooth of the animal is measured.
Ref: Crampton, E. W. (1947). The Journal of Nutrition, 33(5), 491--504. doi: 10.1093/jn/33.5.491.

```{r}
head(ToothGrowth)
dim(ToothGrowth)
```

There are 60 rows and 3 columns.
`len` is tooth length. It will be our dependent variable.
There are 2 kinds of `supp`: "VC" or "OJ".
There are multiple levels of dose, which is recorded as a numeric variable in the table. 

This experiment is called a multifactorial design because it has two independent variables of interest: `supp` and `dose`.
`supp` is one factor; `dose` is another factor. We have two factors of interest, thus multifactorial.

# Why would one want to do a multifactorial experiment?
The multifactorial design is popular because it is efficient.
When we have multiple independent variables of interest,
it allows us to compare multiple combinations of independent variables at the same time.

Let's use the ToothGrowth data as an example.
There are two supplements and three doses of each.
2 * 3 = 6 combinations.
The multifactorial design will allow us to compare all 6 combinations at the same time.

## Be thoughtful about the purpose and budget of your project. 
For multifactorial experiments, the number of samples (and thus the cost) increases quickly! 
Let's say we have `k` independent factors, each with `M` levels and `n` replicates each. 
The total number of samples will be: `M1 * M2 * M3 * ... Mk * n`. 
For every new factor that we include, we are multiplying by the level of the new factor, so
the number of samples can get out of hand quickly. 

When designing a multifactorial experiment, think about: 

1. Which factors should be included to best fulfill the goals of the project? 
2. How many levels in each factor are relevant to the project?  
3. How many factors and levels can I afford to include given the budget? 

# What is interaction among factors?
A very important concept of the multifactorial design is interaction.
Interaction stands for "the effect of one factor depends on the other factor". 
To understand this, let's visualize the data first.
```{r}
ToothGrowth %>% 
  ggplot(aes(x = dose, y = len)) +
  stat_summary(geom = "line",  
               fun.data = mean_se,  
               aes(group = supp, color = supp),  
               size = 1.2) + 
  stat_summary(geom = "linerange",   
               fun.data = mean_se, 
               aes(group = supp,
                   color = supp), 
               alpha = 0.5) +   
  ggbeeswarm::geom_quasirandom(aes(color = supp)) +    
  scale_color_manual(values = c("grey20","tomato1")) +
  scale_x_continuous(breaks = c(0.5, 1, 2)) +  
  labs(x = "dose",              
       y = "teeth length",
       color = "supplement") +
  theme_classic()
```
You can see that as the dose of either supplement increases, the teeth length also increases.
For both supplements, the higher dose led to longer teeth.
However, if you look carefully at the lines, you will notice that the two curves are not exactly parallel.
While teeth are longer for OJ than VC in lower doses, there is no difference in dose = 2.
And **THAT** is interaction. The effect of supplement depends on the dose.
The effect of one factor (supplement) depends on the other factor (dose). 
In the graphical sense, it means the trends are not all parallel to each other.
When you set up a multifactorial design, you will most certainly encounter interactions.

# How do you analyze a multifactorial experiment? 
You will analyze a multifactorial experiment like a randomized block experiment,
except that you need to consider interactions now.

Let's fill out the table.

| source | factor | levels |
|:------:|:------:|:------:|
|  supp  | treat  |   2    |
|  dose  | treat  |   3    |
| animal | OU, EU |   60   |

We have supp and dose as our treatments.
They have 2 and 3 levels, respectively.
We have 60 observations of animal (each dot on the graph is an animal).
In this case, the animal is both observational units and experimental units.

Last time, we learned that we should include all sources of variation except OU.
So that gives `len ~ supp + dose`.

But how do I include the interaction between supp and dose in my linear model? 

## Setting up linear model for multifactorial design
To include an interaction in a linear model,
you use the `factor1:factor2` syntax.   

That is `len ~ supp + dose + supp:dose`

A simplified syntax is `len ~ supp * dose`.
The `*` symbol means `+ and interaction`, that is `supp * dose` = `supp + dose + supp:dose`

Oftentimes people refer to the independent variables as the "main effects".
So supp and dose are your main effects.

When you set up your model, you type in the main effects first, then the interactions.

*WARNING*: The dose is recorded as a numeric variable in the dataset.
In a multifactorial experiment, we need to set it as a factor.
```{r}
ToothGrowth_new <- ToothGrowth %>% 
  mutate(dose.f = as.factor(dose))   
```

We can pull out the summary statistics first:
In this case we will group by both supp and dose.f, the two factors.
```{r}
ToothGrowth_new %>% 
  group_by(supp, dose.f) %>% 
  summarise(mean = mean(len),
            var = var(len),
            sd = sd(len),
            n = n()) %>% 
  ungroup()
```
The variances are all within an order of magnitude, so it's probably going to be okay. 

```{r}
model_teeth <- lm(len ~ supp + dose.f + supp:dose.f, data = ToothGrowth_new) 
```
You can also do `len ~ supp * dose.f`. That gives you the same thing.

Note that you can also set it up as `len ~ dose.f * supp`.
Technically, there is a minute difference between the two models,
but oftentimes it makes no difference to your conclusions at the end of the day.

Let's check the assumptions of ANOVA first.
Normality?
```{r}
plot(model_teeth, which = 2)
```
Pretty good!
Next, equal variance?
```{r}
plot(model_teeth, which = c(1, 3))
```
Looks okay to me.
We are all clear to do ANOVA.

## ANOVA
```{r}
anova(model_teeth)
```

1.  supp has a F >> 1. That means the means of supp are different.

2.  dose.f has F >> 1 but also larger than the F value of supp. This means not only are the means of dose.f different, but also dose.f explains a larger proportion of variation in the data than supp.

3.  Lastly, the interaction supp:dose.f has a F value significantly > 1 (p = 0.02 < 0.05).
    This means the effect of supp depends on the dose, consistent with what the graph showed us.

Now we can go ahead and find out which supp:dose combination gave us the longest teeth.

# Setting up Tukey test correctly for multifactorial experiments
When you do a Tukey test on a multifactorial experiment, you need to be VERY THOUGHTFUL.
Ask yourself what is the comparison that I am actually interested in?

In this example:

* Am I interested in the effect of supp at each dose level? 
* Am I interested in the effect of dose for each supp?

Depending on the question of interest, you will need a different Tukey test layout.
The choice of analysis depends on the researcher and their interests.

## Test effect of supp at each dose level
Let's say we are interested in the effect of supp at each dose level.
To specify that in the `emmeans()` command, we will use the `~ supp | dose` syntax.
This means comparing across supp at each dose level.
```{r}
estimate_per_dose <- emmeans(model_teeth,pairwise ~ supp | dose.f) 
cld(estimate_per_dose$emmeans, Letters = letters)
```
This tells us that at dose = 0.5 or 1, OJ gives longer teeth than VC.
However, at dose = 2, OJ and VC are not significantly different. 

## Test the effect of dose at each supp
Now let's say we are interested in the effect of different doses for each supp.
How should we set up the `emmeans()` command?

You'll use `~ dose.f | supp` now.
This means comparing across dose at each supp level.
```{r}
estimate_per_supp <- emmeans(model_teeth,pairwise ~ dose.f | supp) 
cld(estimate_per_supp$emmeans, Letters = letters)
```
This tells us that

1.  For OJ, dose 0.5 is significantly lower than dose 1 and 2.
    Dose 2 is not significantly different from dose 1.

2.  For VC, as the dose increases, each dose is significantly different from the previous.

Hopefully you can see that depending on how we set up `emmeans()`, we get different inferences.
Before you set up a Tukey test for a multifactorial experiment, do think about what you are actually interested in.
 
# In class activity
## Q1 
Load "S8_activity1.csv" from the Data folder.  
In this experiment, we are looking at the yield of two varieties of potatoes across different nitrogen input.
Potatoes from one of the two varieties were randomly assigned one of the 5 nitrogen treatments.
At the end of the experiment, potatoes were harvested and weighed.
(Data from UC Davis Plant Sciences course PLS205, Winter 2017)

1. Visualize the data. 
Hint: Are there any numerically recorded variables that need to be changed into characters? 

2. Set up a linear model. Use an experimental design table to help you. 
Hint: What main effects need to be included? What interaction needs to be included?  

3. Check assumptions of linear models. 

4. Run ANOVA and interpret the ANOVA table. 

5. Set up the Tukey test. 
Let's say we are interested in knowing which variety performs better at each nitrogen input.
How should you set up the Tukey test?

