---
title: "S9 in class activity"
author: "Chenxin Li"
date: "2026-06-24"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In this unit, we will go over analyses regarding:

* Association between two categorical variables (Fisher Exact test)
* Proportions (Chi-squared test)
* Enrichment or underrepresentation (Hypergeometric test)

# Packages
No external packages are required today. 

# Testing association between two categorical variables 
If we want to test the association between 2 continuous variables, 
we will do a correlation test. 
However, if we want to test the association between 2 categorical variables, 
we will need something different. 
A general approach is using the Fisher Exact test. 

Here are the assumptions: 

1. Variables being tested are categorical 
2. Categories must be mutually exclusive 
3. Observations are independent 

Let's start with an example. 
A person claims that they can tell the difference between a can of Coke vs a can of Pepsi. 
They were given samples of each and blind tasted them. 
The results are the following: 

|               | actually Coke | actually Pepsi |
|:-------------:|:-------------:|:--------------:|
| claimed Coke  |      9        |        3       |
| claimed Pepsi |      1        |        7       |

The two categorical variables here are: 

1. The true identity of the drink: Coke or Pepsi.
2. What the person thinks the drink is: Coke or Pepsi. 

If there is a strong association between these two variables, 
i.e., when the person thinks the drink is Coke, it's actually also Coke, and vice versa,  
we can conclude that this person can distinguish between these two drinks. 

Given the results, how well can this person distinguish these two drinks? 
The table presented above is called a contingency table. 
In a contingency table, you put one categorical variable as the rows,
and the other as the columns: 

|               | actually Coke | actually Pepsi |
|:-------------:|:-------------:|:--------------:|
| claimed Coke  |      A        |        B       |
| claimed Pepsi |      C        |        D       | 

One metric you can calculate from a contingency table is the odds ratio (OR). 
It's the product of one diagonal over the other, or OR = (A * D)/(B * C). 
In this example, it would be (9 * 7) / (1 * 3) = 21, 
which can be understood as this person is 21 times more likely to call things right than call things wrong. 
Without doing any hypothesis testing, we already have some idea based on the OR. 
The OR has a null hypothesis of 1. 21 is a lot larger than 1,
so we should expect to reject the null hypothesis. 

We will use `fisher.test()` to perform a Fisher Exact test by feeding the contingency table into it. 
The easiest way is to combine the two rows into a matrix using `rbind()`, 
or combining the two columns into a matrix using `cbind()`. 
```{r}
fisher.test(x = rbind(
  c(9 ,3),
  c(1, 7)
))
```
You can see the test returned p = 0.01977, which will prompt us to reject the null hypothesis. 
Notice that the OR calculated by the test is similar but not equal to the one we calculated. 
`fisher.test()` uses a more complex formula to calculate OR, 
which returns an OR similar but not equal to the simple AD/BC formula. 

A couple final notes on Fisher Exact test. 

1. The test works on both small and large sample sizes. 
2. When the sample size is small, the test may return a non-significant result even when OR >> 1. This indicates a larger sample size is needed. 

# Testing observed frequencies with known proportions 
The classic example is testing if observed frequencies in a sample 
fits the expected proportions. In this case, you will use Chi-squared test.

Now let's do an example. Let's say I self pollinate a heterozygous plant of genotype `Aa`. 
I expect 1/4`AA`, 1/2 `Aa` and 1/4 `aa` from the offspring. 
In this hypothetical experiment, among 96 F1 offspring, 
I have 32 `AA`, 54 `Aa`, and 10 `aa`.

You can see that 32 is a bit more than 1/4 and 54 is a bit more than 1/2, 
which means 10 is a bit less than 1/4. 
How do you test if the observed frequencies fit the expected ratio?

You will do a Chi-square test for goodness of fit.
```{r}
chisq.test(x = c(32, 54, 10), 
           p = c(0.25, 0.5, 0.25))
```
In `chisq.test()`, you provide two arguments. 
First is the observed frequencies. In this case, a vector of three observations, 32, 54 and 10. 
The second argument is the expected proportions, which is 0.25, 0.5 and 0.25.

We have Chi-squared = 11.583, and p value = 0.003. 
The p value is computed from the Chi-squared distribution. 
In this case, we have 2 degrees of freedom, which is the number of categories - 1. 
We have 3 categories, `AA`, `Aa` and `aa`, so df = 3 - 1 = 2.
The null hypothesis of Chi-squared is 0.
A Chi-squared value of 0 would mean the observed data fit the expected proportions perfectly.  

So what do our results (p = 0.003) mean? 
It means that there is more `AA` and `Aa` than expected, and less `aa` than expected. 
In a real world example, this could mean a lethality in the `aa` genotype. 

You noticed that here I used 96 plants, which was a lot. 
What will happen if there were less plants? 
Let's say in a hypothetical experiment, I only have 20 F1 offspring from selfing the `Aa` parent. 
I got 7 `AA`, 11 `Aa`, and 2 `aa`, which is similar in proportion to the n = 96 example above. 
If I put these numbers into a Chi-squared test, what will happen?
```{r}
chisq.test(x = c(7, 11, 2), p = c(0.25, 0.5, 0.25))
```
Now p = 0.26, which is larger than 0.05 and you would fail to reject the null hypothesis. 
*Be VERY CAREFUL* when you interpret large p values! 
Note that a p  > 0.05 does not prove the null hypothesis. It simply fails to reject it. 
The true proportions may deviate from the expected ratio, but you didn't have enough observations (power) to detect it. 

Final notes for Chi-squared test: 

1. As a rule of thumb, to perform a valid Chi-squared test, you need many observations, ideally  > 100.
2. The Chi-squared test requires expected proportions. If the expected proportions are unknown, we can't do the test. 

# Testing enrichment or underrepresentation 
Let's say if we have a subset or a sample from a population, and you want to test if 
the sample was representative of the population, 
or if certain entities are enriched or underrepresented, what should you do? 

Consider this simple experiment: 

1. There are 10 white balls and 90 black balls in a urn. 
2. You drew a sample (without replacement) of 11 balls containing 6 white and 5 black balls. 
3. You wonder: Are white balls enriched in my sample? 

## Testing enrichment using a contingency table 
You can approach this problem using a contingency table. 
In this case, the 2 categorical variables are: 

1. Was the ball drawn: yes or no
2. What color was it: white or black

|           | white    | black  |
|:---------:|:--------:|:------:|
| drawn     |     6    |    5   |
| not drawn |     4    |    85  |

The odds ratio will be (6 * 85) / (4 * 5) = 25.5, which is a lot greater than 1. 
Thus, you can infer that there is an association between a ball that is drawn and white. 
```{r}
fisher.test(rbind(
  c(6,5),
  c(4,85)
))
```
The Fisher Exact test will return a p value of 6.7e-5, which is quite significant, as we expected. 

## Testing enrichment using the hypergeometric test 
An equally valid approach is using the hypergeometric test,
which can be used to test for enrichment or underrepresentation. 
It takes 4 parameters: `q`, `m`, `n`, `k`. 

Imagine your are randomly drawing a certain number of balls from an urn (or bag) without replacement. 
In the urn, there are white balls and black balls. There are no other colors.

-   q is number of white balls you drew.
-   m is total number of white balls in the urn.
-   n is total number of black balls (or not white balls) in the urn.
-   k is total number of balls you drew.

Coming back to our example earlier: 

-   q = 6 (white balls drawn)
-   m = 10 (total white balls) 
-   n = 90 (total not white balls) 
-   k = 11 (total balls drawn) 

The function is `phyper()`.
```{r}
phyper(
  q = 6, 
  m = 10,
  n = 90,
  k = 11,
  lower.tail = F
)
```
Note that I used the `lower.tail = FALSE` option. 

* If you are testing for enrichment, use `lower.tail = FALSE`.
* If you are testing for underrepresentation, use `lower.tail = TRUE`. 

The p-value is 2.2e-6, which is consistent with what Fisher exact test produced. 

A final note for the hypergeometric test is that it works on both large and small sample sizes. 

# In class activity 
## Q1
You are working on a specialized cell type called cell type X.  

* According to the literature, 4% of genes are cell type X markers (most highly expressed in cell type X). 
* You perform a perturbation experiment and found that 796 genes were up-regulated. 
* Among them, 147 genes were cell type X markers. 

Perform a statistical test to test whether there are significantly more cell type X markers among the up-regulated genes.   
Hint: Based on the information given, which statistical test can you use? 

## Q2 
You are working on a metabolic pathway that is only expressed in a specialized cell type, cell type X. 

* You have discovered 31 transcription factors (TFs) that activate or repress the metabolic pathway. 
* Among the 31 TFs, 13 were activators, and 18 were repressors. 
* Among the 13 activators, 12 were expressed in cell type X. 
* Among the 18 repressors, 2 were expressed in cell type X. 

1. Set up a contingency table for the data presented above. 
2. Perform a statistical test to test whether there is an association where activators tend to be expressed in cell type X, and repressors tend to not be expressed in cell type X. 
 
## Q3
Consider the following: 

1. In an RNA-seq experiment, you detected 12000 expressed genes. 
2. Among the expressed genes, 2000 of them were annotated to be "growth related".
3. You have a list of 130 up-regulated genes responding to a stimulus. 
4. 60 of the up-regulated genes are annotated as "growth related".  

Were growth related genes enriched among the up-regulated genes in response to the stimulus? 

## Q4 
You wrote an algorithm to classify RNA-seq samples as "healthy" or "pathological" based on the expression of genes known to be involved in a certain disease. 

1. You tested the algorithm on 30 new samples. 
2. 20 samples were detected as "healthy" and originated from a healthy donor. 
3. 5 samples were detected as "pathological" but in fact originated from a healthy donor.
4. 1 sample was detected as "healthy" but in fact came from an unhealthy donor. 

How well did the classification algorithm perform?  
Hint: What can be used as metrics for a well-performing classification algorithm? 




